Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-06-08 documents/ms.tex
finishing off - still need to read over once more

@@

\begin{abstract}
Variability in the light curves of spotted, rotating stars is often
    non-sinusoidal and quasi-periodic --- spots move on the stellar surface
    and have finite lifetimes, causing stellar flux variations to slowly shift
    in phase.
A strictly periodic sinusoid therefore cannot accurately model a rotationally
    modulated stellar light curve.
Physical models of stellar surfaces have many drawbacks preventing effective
    inference, such as highly degenerate or high-dimensional parameter spaces.
In this work, we test an appropriate {\it effective} model: a Gaussian
    Process with a quasi-periodic covariance kernel function.
This highly flexible model allows sampling of the posterior Probability
    Density Function (PDF) of the periodic parameter, marginalising over the
    other kernel hyperparameters using a Markov Chain Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
    \naigrain\ simulated stellar light curves, demonstrating that the Gaussian
    process method produces periods that are slightly more accurate than both
    a sine-fitting periodogram and an autocorrelation function method.
We also demonstrate that it works effectively on real data, by inferring
    rotation periods for \nkoimcq\ \Kepler\ stars with previously measured
    periods.
Because this method delivers posterior PDFs, it will enable hierarchical
    studies involving stellar rotation, particularly those involving
    population modelling, such as inferring stellar ages, obliquities in
    exoplanet systems, or characterising star-planet interactions.
The code used to implement this new method is available online.
% The ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

In this work, we test an {\it effective} model for rotationally modulated
stellar light curves which captures the salient behaviour but is not
physically motivated --- although some parameters may indeed be {\it
interpreted} as physical ones.
An ideal effective model should have a small number of non-degenerate
parameters and be flexible enough to perfectly capture non-sinusoidal and QP
behaviour.
A Gaussian process (GP) model fulfills these requirements. We thus use a GP
as the generative model at the core of a method to probabilistically
infer accurate and precise stellar rotation periods.
This enables us to estimate the posterior PDF of the rotation period, thereby
producing a justified estimate of its uncertainty.
{\bf Using a quasi-periodic GP model to infer a rotation period is not a new idea.
We used this method in \citet{Vanderburg2015} to measure the rotation period
of an exoplanet host.
Previously to that, \citet{Haywood2014} used the quasi-periodic GP model
to disentangle the correlated noise produced by stellar activity and rotation
from the radial velocity signature of an exoplanet.
In addition, \citet{Littlefair2017} use a quasi-periodic GP to establish the
quasi-periodicity of variability in the light curves of brown dwarfs.
What we present here is the first test of the GP method, placed alongside
alternatives.
We also provide code allowing others to easily apply this technique
themselves.}

@@

We compare our results to those of the Tel Aviv team in \citet{Aigrain2015}
who use the ACF method described in \citet{Mcquillan2014} and
\citet{Aigrain2015}.
{\bf This method involves the calculation of an autocorrelation function, a
measure of the self-similarity of the light curve over a range of lags, which
is then smoothed with a Gaussian kernel.
Periodic variability in the light curve produces a peak at the period of the
signal in the ACF and a decaying sequence of peaks at integer multiples of the
period.
\citet{McQuillan2014} first pre-processed the light curves by removing very
long-term variations.
They then calculated an ACF for each star and measured the positions of up to
the first four peaks in the ACF.
They fit a straight line to the periods of these peaks as a function of their
integer multiple and adopted the gradient of that line as the rotation
period.
The uncertainty on the ACF period is the uncertainty on the best-fit slope.
}
Figure \ref{fig:compare_acf} shows ACF-measured versus true rotation
periods, with the $2n$ and $\frac{1}{2}n$ harmonic lines as dashed lines.
The injected and recovered rotation periods generally agree well: the MRD of
their periods, relative to the true rotation periods, is \percenttelavivMAD\%
--- slightly smaller than the results produced by the GP method.
However, their results contain a larger number of extreme outliers --- they
predict periods that are significantly different from the true period in more
cases than the GP method.
This is revealed by comparing the RMS of the two sets of results since the RMS
is more sensitive to outliers.
The RMS of the Tel Aviv method results is \telavivRMS\ days and the RMS of GP
method results is \gpRMS\ days.
% Another negative feature of the ACF method is revealed by examining the points
% clustered close to the 1:1 line: more fall slightly below than above it, \ie\
% rotation periods are systematically underestimated.
% This stems from peak position measurements in the ACF method.
% ACFs of stellar light curves have similar functional forms to the QP kernel
% function in Equation \ref{eq:QP}: periodic functions added to decaying
% exponentials.
% In such functions the peak positions can be shifted towards the left (towards
% shorter periods), because the decaying exponential raises the left side of
% each peak more than the right.
% It is possible to model this effect; however, standard practice
% simply measures the peak position without taking it into account.

@@

{\bf We first applied a high-pass filter to the light curves, removing
long-term trends that could produce large peaks at long periods in the
periodogram.
We used the 3rd order Butterworth filter in {\tt scipy} python with a cut-off
of 35 days, attenuating signals with periods greater than this threshold.
This filter was applied in frequency space, rather than period space, and full
attenuation was attained at zero frequency.
In other words, the level of attenuation increased smoothly with period, for
periods above 35 days but no periods were fully suppressed.
Experimentally, we found the 35 day cut off removed the largest number of long
period outliers while preserving shorter period signals.
We also tried 50, 45, 40 and 30 day cut-offs but the 35 day cut-off value
maximised the RMS and MAD of the LS periodogram results.}
For each simulated light curve, we compute a LS periodogram\footnote{LS
periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.} over a
grid of 10,000 periods, evenly spaced in frequency, between 1 and 100 days.
We adopt the period of the highest peak in the periodogram as the measured
rotation period.
{\bf The uncertainties on the rotation periods were calculated using the
following equation for the standard deviation of the frequency
\citep{Horne1986, Kovacs1981}:
\begin{equation}
    \sigma_{\nu} = \frac{3\pi\sigma_N}{2N_O^{1/2}TA},
\end{equation}
where $A$ is the amplitude of the signal of highest power, $\sigma_N$ is the
variance of the time series, with the signal of highest power removed, $N_O$
is the number of variations and $T$ is the timespan of the data.
These formal uncertainties are only valid in the case that the noise is white,
the data are evenly sampled and there is only one signal present.
Since there are multiple signals present in these light curves, this formal
uncertainty is an underestimate of the true uncertainty.
}
Figure \ref{fig:pgram_compare} shows the resulting recovered rotation periods
as a function of true period.
{\bf The MRD of the periodogram-recovered periods is \percentpgramMAD\%,
slightly worse than the ACF and GP methods (see table \ref{tab:MADs} for
a side-by-side comparison with the other methods).
The filter suppresses signals with periods greater than 35 days but doesn't
eliminate them altogether, thus allowing high amplitude frequencies to be
detectable.
For this reason, some erroneous long periods are recovered by the periodogram
method.
These results still show a marked improvement in comparison to periodograms
calculated from unfiltered data.
Simply adopting the highest peak of a periodogram calculated from the
unfiltered simulated data resulted in an RMS of \oldpgramRMS\ and relative MAD
of \oldpercentpgramMAD.}
% The LS periodogram method drastically overestimates some rotation periods
% because long-term trends unrelated to rotation still produced excess power at
% long periods, despite our attempts to remove these trends via filtering.
% {\bf Filtering a light curve before attempting to measure a rotation period is
% not always advisable because it may remove real signals from the data.
% The simulated light curves used in this experiment had rotation periods of up
% to 50 days, and thus applying a high pass filter with a 35 day cut off left
% most of these periods unaffected.
% Of course, this would not be the case if there were many slower rotators in
% this sample.
% In some cases a signal greater than 35 days was recovered, sometimes correctly
% and sometimes erroneously.
% }
% Filtering is not analogous to using priors in Bayesian inference as filtering
% completely removes signals; if a periodic signal is filtered out there is no
% hope of recovering it.
% Using priors that favour shorter periods over longer ones still allows for the
% detection of long periods if they are strongly supported by the data.}

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, demonstrating that this
method works not only on simulated data, but also on real data---with the
caveat that for any particular data set, some care is needed regarding the
setting the priors and ranges for the GP hyperparameters.
% Notably, the GP method recovers periods systematically slightly larger than
% \citet{Mcquillan2013}---a likely consequence of correcting for the ACF peak
% measurement bias discussed in Section \ref{sec:acf}.

@@

{\bf The majority of long period outliers seen in the periodogram comparison
figure (figure \ref{fig:compare_pgram} results are produced when the rotation
period signal is highly non-sinusoidal.
This can be understood intuitively as the LS periodogram fits a series of sine
waves to the data; if the data are not sinusoidal, a sine wave will not
produce a good fit.
It is difficult to assess the reason for the ACF incorrect recoveries because
we do not have the individual ACFs computed by the \citet{Aigrain2015} Tel
Aviv team, however it seems that the ACF method is more sensitive than the LS
periodogram method to noise: this particular implementation of the LS
periodogram method is capable of detecting periodic signals at lower
amplitudes that the ACF.
We do not suggest that the LS periodogram method is better than the ACF method
in general.
It was made clear in \citet{Aigrain2015} that the performance of each of these
rotation period detection algorithms depends sensitively on the exact
implementations.
The exact steps performed in the pre-processing of the light curves and the
post-processing of the periodogram or ACF seems to hugely affect the results.
This is also true for the GP method.
Using a different MCMC sampler, different priors, different subsampling, or a
different kernel would all alter the results.
Clearly, none of these rotation period measurement methods is perfect and they
all require a lot of background engineering and tuning.
This is simply due to inherent difficulty of inferring periods, or any
parameter that produces a highly multimodal probability distribution, from
noisy data.}

@@

In addition, we are continuing to develop several other aspects of this GP
method:
\begin{itemize}
    \item{{\bf To use an updated method for calculating the GP likelihood and
    drastically improve computational efficiency.
        {\tt celerite}, a new method for rapid computation of the 'solve' and
        determinant calculations of covariance matrices necessary for GP
        inference has recently been developed \citep{Foremanmackey2017}.
        {\tt celerite} scales with the number of data points, $\mathcal{O}(N)$
        rather than the $N\log(N)^2$ scaling of the {\tt HODLR} algorithm used
        in this analysis.
        It requires a slightly different functional form for the kernel as the
        increased speed is obtained by using mathematical properties of
        exponential functions.
        Kernels must be written as the sums of exponentials in order to be
        implemented in {\tt celerite}, however a close approximation to the
        kernel function used here is demonstrated to recover rotation periods
        in \citet{Foremanmackey2017}.}}
\item{To investigate more physically motivated priors.
        In this paper we have only explored the physical interpretation of one
        parameter in our kernel function (P), which we related to stellar
        rotation periods, but others also warrant physically motivated priors.
        For example, the harmonic complexity parameter, $\Gamma$, defines the
        typical number of inflection points per function period; suitable
        priors may be informed by the typical properties of observed light
        curves.
        For the evolution time-scale, $l$,  it is sensible to enforce $l > P$,
        such that functions that have clear periodicity.
        Intuitively, if $l < P$, functions may evolve significantly over time
        scales shorter than one period, so a function $f(t+P)$ would look
        significantly different to $f(t)$, in which case any claims of
        periodicity become dubious.
        More stringent criteria may be suggested by stellar astrophysical
        considerations, e.g. knowledge of typical active region evolution
        time-scales or persistence lifetimes (e.g. $l\sim2P$).
        Finally, the amplitude hyper-parameter $A$ may be related to active
        region coverage.
        Unfortunately, because these hyper-parameters control the covariance
        structure of functions, their interpretation is not always
        straightforward; interpretation is further complicated by significant
        degeneracies between the hyperparameters.
        Therefore we defer such considerations, along with strategies for
        efficient optimization or marginalization of hyperparameters, to a
        separate paper (currently in prep.).}
\item{To build in a noise model for \kepler\ data.
        Because it is a {\it generative} model of the data, the GP method
        models the rotation period at the same time as systematic noise
        features.  One can then marginalise over the parameters of the noise
        model.
        This approach would be extremely advantageous for \kepler\ data since
        long-term trends are often removed by the \kepler\ detrending
        pipeline.  Marginalising over the noise model at the same time as
        inferring the parameters of interest will insure that the periodic
        signal is preserved.}
\end{itemize}

@@

The main aim of this work was to develop a probabilistic rotation period
inference method.
Probabilistic periods are necessary for hierarchical Bayesian inference,
particularly when performing population analysis.
This new GP method is capable of generating a catalogue of probabilistic
rotation periods which could, for example, reveal the period distribution
(and therefore potentially the age distribution) of stars in the Milky Way.
There is still room for improvement since the uncertainty estimates are not
yet truly representative and it is still only an `effective' model, not a
physical one.
However, it is probabilistic and provides more accurate periods than
alternative methods.
{\bf Although GP inference is slow as it requires a `solve' and a matrix
determinant calculation at every step of an MCMC, a new GP solver, {\tt
celerite} \citep{Foremanmackey2017} shows promise for increasing the
computional speed of GP rotation period inference by several orders of
magnitude.
We intend to integrate {\tt celerite} into our code in the near future.}
The code used in this project is available at
\url{https://github.com/RuthAngus/GProtation/}.

>>>>

\newcommand{\nkoi}{1132}

@@

\begin{abstract}
Variability in the light curves of spotted, rotating stars is often
    non-sinusoidal and quasi-periodic --- spots move on the stellar surface
    and have finite lifetimes, causing stellar flux variations to slowly shift
    in phase.
A strictly periodic sinusoid therefore cannot accurately model a rotationally
    modulated stellar light curve.
Physical models of stellar surfaces have many drawbacks preventing effective
    inference, such as highly degenerate or high-dimensional parameter spaces.
In this work, we test an appropriate {\it effective} model: a Gaussian
    Process with a quasi-periodic covariance kernel function.
This highly flexible model allows sampling of the posterior Probability
    Density Function of the periodic parameter, marginalising over the
    other kernel hyperparameters using a Markov Chain Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
    \naigrain\ simulated stellar light curves, demonstrating that the Gaussian
    process method produces periods that are slightly more accurate than both
    a sine-fitting periodogram and an autocorrelation function method.
We also demonstrate that it works effectively on real data, by inferring
    rotation periods for \nkoimcq\ \Kepler\ stars with previously measured
    periods.
Because this method delivers posterior probability density functions, it will
    enable hierarchical studies involving stellar rotation, particularly those
    involving population modelling, such as inferring stellar ages,
    obliquities in exoplanet systems, or characterising star-planet
    interactions.
    {\bf We provide a table of rotation periods for \nkoi\ \Kepler\ objects of
    interest and full posterior probability density function samples.}
    The code used to implement this new method is available online.
% The ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

In this work, we test an {\it effective} model for rotationally modulated
stellar light curves which captures the salient behaviour but is not
physically motivated --- although some parameters may indeed be {\it
interpreted} as physical ones.
An ideal effective model should have a small number of non-degenerate
parameters and be flexible enough to perfectly capture non-sinusoidal and QP
behaviour.
A Gaussian process (GP) model fulfills these requirements. We thus use a GP
as the generative model at the core of a method to probabilistically
infer accurate and precise stellar rotation periods.
This enables us to estimate the posterior PDF of the rotation period, thereby
producing a justified estimate of its uncertainty.
{\bf Using a quasi-periodic GP model to infer a rotation period is not a new idea.
We used this method in \citet{Vanderburg2015} to measure the rotation period
of an exoplanet host.
Previously to that, \citet{Haywood2014} used the quasi-periodic GP model
to disentangle the correlated noise produced by stellar activity and rotation
from the radial velocity signature of an exoplanet.
In addition, \citet{Littlefair2017} use a quasi-periodic GP to establish the
quasi-periodicity of variability in the light curves of brown dwarfs.
What we present here is a test of the GP method and a comparison with
alternative rotation period measurement methods.
We also provide code allowing others to easily apply this technique
themselves.}

@@

We compare our results to those of the Tel Aviv team in \citet{Aigrain2015}
who use the ACF method described in \citet{Mcquillan2014} and
\citet{Aigrain2015}.
{\bf This method involves the calculation of an autocorrelation function, a
measure of the self-similarity of the light curve over a range of lags, which
is then smoothed with a Gaussian kernel.
Periodic variability in the light curve produces a peak in the ACF at the
period of the signal and a decaying sequence of peaks at integer multiples of
the period.
\citet{Mcquillan2014} first pre-processed the light curves by removing very
long-term variations.
They then calculated an ACF for each star and measured the positions of up to
the first four peaks in the ACF.
They fit a straight line to the periods of these peaks as a function of their
integer multiple and adopted the gradient of that line as the rotation
period.
The uncertainty on the ACF period is the uncertainty on the best-fit slope.
}
Figure \ref{fig:compare_acf} shows ACF-measured versus true rotation
periods, with the $2n$ and $\frac{1}{2}n$ harmonic lines as dashed lines.
The injected and recovered rotation periods generally agree well: the MRD of
their periods, relative to the true rotation periods, is \percenttelavivMAD\%
--- slightly smaller than the results produced by the GP method.
However, their results contain a larger number of extreme outliers --- they
predict periods that are significantly different from the true period in more
cases than the GP method.
This is revealed by comparing the RMS of the two sets of results since the RMS
is more sensitive to outliers.
The RMS of the Tel Aviv method results is \telavivRMS\ days and the RMS of GP
method results is \gpRMS\ days.
% Another negative feature of the ACF method is revealed by examining the points
% clustered close to the 1:1 line: more fall slightly below than above it, \ie\
% rotation periods are systematically underestimated.
% This stems from peak position measurements in the ACF method.
% ACFs of stellar light curves have similar functional forms to the QP kernel
% function in Equation \ref{eq:QP}: periodic functions added to decaying
% exponentials.
% In such functions the peak positions can be shifted towards the left (towards
% shorter periods), because the decaying exponential raises the left side of
% each peak more than the right.
% It is possible to model this effect; however, standard practice
% simply measures the peak position without taking it into account.

@@

{\bf We first applied a high-pass filter to the light curves, removing
long-term trends that could produce large peaks at long periods in the
periodogram.
We used the 3rd order Butterworth filter in {\tt scipy} with a cut-off
of 35 days, attenuating signals with periods greater than this threshold.
This filter was applied in frequency space, rather than period space, and full
attenuation was attained at zero frequency.
In other words, the level of attenuation increased smoothly with period, for
periods above 35 days but no periods were fully suppressed.
Experimentally, we found the 35 day cut off removed the largest number of long
period outliers while preserving shorter period signals.
We also tried 50, 45, 40 and 30 day cut-offs but the 35 day cut-off value
maximised the RMS and MAD of the LS periodogram results.}
For each simulated light curve, we computed a LS periodogram\footnote{LS
periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.} over a
grid of 10,000 periods, evenly spaced in frequency, between 1 and 100 days.
We adopted the period of the highest peak in the periodogram as the measured
rotation period.
{\bf The uncertainties on the rotation periods were calculated using the
following equation for the standard deviation of the frequency
\citep{Horne1986, Kovacs1981}:
\begin{equation}
    \sigma_{\nu} = \frac{3\pi\sigma_N}{2N^{1/2}TA},
\end{equation}
where $A$ is the amplitude of the signal of highest power, $\sigma_N$ is the
variance of the time series, with the signal of highest power removed, $N$
is the number of observations and $T$ is the timespan of the data.
These formal uncertainties are only valid in the case that the noise is white,
the data are evenly sampled and there is only one signal present.
Since there are multiple signals present in these light curves, this formal
uncertainty is an underestimate of the true uncertainty.
}
Figure \ref{fig:pgram_compare} shows the resulting recovered rotation periods
as a function of true period.
{\bf The MRD of the periodogram-recovered periods is \percentpgramMAD\%,
slightly worse than the ACF and GP methods (see table \ref{tab:MADs} for
a side-by-side comparison with the other methods).
The filter suppresses signals with periods greater than 35 days but doesn't
eliminate them altogether, thus allowing high amplitude frequencies to be
detectable.
For this reason, some erroneous long periods are recovered by the periodogram
method.
These results still show a marked improvement in comparison to periodograms
calculated from unfiltered data.
Simply adopting the highest peak of a periodogram calculated from the {\it
unfiltered} simulated data resulted in an RMS of \oldpgramRMS\ and relative
MAD of \oldpercentpgramMAD.}
% The LS periodogram method drastically overestimates some rotation periods
% because long-term trends unrelated to rotation still produced excess power
% at long periods, despite our attempts to remove these trends via filtering.
% {\bf Filtering a light curve before attempting to measure a rotation period is
% not always advisable because it may remove real signals from the data.
% The simulated light curves used in this experiment had rotation periods of up
% to 50 days, and thus applying a high pass filter with a 35 day cut off left
% most of these periods unaffected.
% Of course, this would not be the case if there were many slower rotators in
% this sample.
% In some cases a signal greater than 35 days was recovered, sometimes correctly
% and sometimes erroneously.
% }
% Filtering is not analogous to using priors in Bayesian inference as filtering
% completely removes signals; if a periodic signal is filtered out there is no
% hope of recovering it.
% Using priors that favour shorter periods over longer ones still allows for the
% detection of long periods if they are strongly supported by the data.}

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, demonstrating that this
method works not only on simulated data, but also on real data---with the
caveat that for any particular data set, some care is needed regarding the
setting the priors and ranges for the GP hyperparameters.
% Notably, the GP method recovers periods systematically slightly larger than
% \citet{Mcquillan2013}---a likely consequence of correcting for the ACF peak
% measurement bias discussed in Section \ref{sec:acf}.
{\bf The new GP rotation periods calculated for \nkoi\ KOIs are provided in an
online table accompanying this paper.
The first few lines of this table is shown below
(table \ref{tab:koi_periods}).
Samples of the rotation period posterior PDFs are available online:
\url{https://doi.org/10.5281/zenodo.804440}}

\begin{table*}
\begin{center}
    \caption{Rotation periods and \Kepler\ input catalogue (KIC) properties
    for the \Kepler\ objects of interest. The full table is available online.}
\begin{tabular}{lcccc}
    KOI number & Period (days) & KIC T$_{\mathrm{eff}}$ & KIC $\log(g)$ & KIC
    $[$Fe$/$H$]$ \\
    \hline
    1 & 25$\pm1$ & $5126^{+160}_{-138}$ & $4.61^{+0.03}_{-0.1}$ &
    $-0.1\pm0.3$ \\
    2 & 20$\pm1$ & $5850\pm50$ & $4.46\pm0.03$ & $-0.2\pm0.1$ \\
    3 & 28$\pm1$ & $6350\pm80$ & $4.02\pm0.01$ & $0.26\pm0.08$\\
    6 & 23$\pm1$ & $6225^{+114}_{-158}$ & $4.17^{+0.06}_{-0.05}$ &
    $-0.0^{+0.1}_{-0.2}$\\
    7 & 25$\pm1$ & $5543\pm79 $ & $4.08\pm0.01$ & $0.4\pm0.1$ \\
    ... & ... & ... & ... & ... \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:koi_periods}

@@

{\bf The majority of long period outliers seen in the periodogram comparison
figure (figure \ref{fig:pgram_compare}) are produced when the rotation period
signal is highly non-sinusoidal.
This can be understood intuitively as the LS periodogram fits a series of sine
waves to the data; if the data are not sinusoidal, a sine wave will not
produce a good fit.
It is difficult to assess the reason for the ACF incorrect recoveries because
we do not have the individual ACFs computed by the \citet{Aigrain2015} Tel
Aviv team, however it seems that the ACF method is more sensitive than the LS
periodogram method to noise.
This particular implementation of the LS periodogram method is capable of
detecting periodic signals at lower amplitudes that the ACF.
We do not suggest that the LS periodogram method is better than the ACF method
in general.
It was made clear in \citet{Aigrain2015} that the performance of each of these
rotation period detection algorithms depends sensitively on the
implementation technique.
The exact steps performed in the pre-processing of the light curves and the
post-processing of the periodogram or ACF seems to hugely affect the results.
This is also true for the GP method.
Using a different MCMC sampler, different priors, different subsampling, or a
different kernel would all alter the results.
Clearly, none of these rotation period measurement methods is perfect and they
all require a lot of background engineering and tuning.
This is simply due to inherent difficulty of inferring periods, or any
parameter that produces a highly multimodal probability distribution, from
noisy data.}

@@

In addition, we are continuing to develop several other aspects of this GP
method:
\begin{itemize}
    \item{{\bf To use an updated method for calculating the GP likelihood and
    drastically improve computational efficiency.
        {\tt celerite}, a new method for rapid computation of the `solve' and
        determinant calculations of covariance matrices necessary for GP
        inference has recently been developed \citep{Foremanmackey2017}.
        {\tt celerite} scales with the number of data points, $\mathcal{O}(N)$
        rather than the $N\log(N)^2$ scaling of the {\tt HODLR} algorithm used
        in this analysis.
        It requires a slightly different functional form for the kernel as the
        increased speed is obtained by using mathematical properties of
        exponential functions.
        Kernels must be written as the sums of exponentials in order to be
        implemented in {\tt celerite}, however a close approximation to the
        kernel function used here is demonstrated to recover rotation periods
        in \citet{Foremanmackey2017}.}}
\item{To investigate more physically motivated priors.
        In this paper we have only explored the physical interpretation of one
        parameter in our kernel function (P), which we related to stellar
        rotation periods, but others also warrant physically motivated priors.
        For example, the harmonic complexity parameter, $\Gamma$, defines the
        typical number of inflection points per function period; suitable
        priors may be informed by the typical properties of observed light
        curves.
        For the evolution time-scale, $l$,  it is sensible to enforce $l > P$,
        such that functions that have clear periodicity.
        Intuitively, if $l < P$, functions may evolve significantly over time
        scales shorter than one period, so a function $f(t+P)$ would look
        significantly different to $f(t)$, in which case any claims of
        periodicity become dubious.
        More stringent criteria may be suggested by stellar astrophysical
        considerations, e.g. knowledge of typical active region evolution
        time-scales or persistence lifetimes (e.g. $l\sim2P$).
        Finally, the amplitude hyper-parameter $A$ may be related to active
        region coverage.
        Unfortunately, because these hyper-parameters control the covariance
        structure of functions, their interpretation is not always
        straightforward; interpretation is further complicated by significant
        degeneracies between the hyperparameters.
        Therefore we defer such considerations, along with strategies for
        efficient optimization or marginalization of hyperparameters, to a
        separate paper (currently in prep.).}
\item{To build in a noise model for \kepler\ data.
        Because it is a {\it generative} model of the data, the GP method
        models the rotation period at the same time as systematic noise
        features.  One can then marginalise over the parameters of the noise
        model.
        This approach would be extremely advantageous for \kepler\ data since
        long-term trends are often removed by the \kepler\ detrending
        pipeline.  Marginalising over the noise model at the same time as
        inferring the parameters of interest will insure that the periodic
        signal is preserved.}
\end{itemize}

@@

The main aim of this work was to develop a probabilistic rotation period
inference method.
Probabilistic periods are necessary for hierarchical Bayesian inference,
particularly when performing population analysis.
This new GP method is capable of generating a catalogue of probabilistic
rotation periods which could, for example, reveal the period distribution
(and therefore potentially the age distribution) of stars in the Milky Way.
There is still room for improvement since the uncertainty estimates are not
yet truly representative and it is still only an `effective' model, not a
physical one.
However, it is probabilistic and provides more accurate periods than
alternative methods.
{\bf Although GP inference is slow as it requires a `solve' and a matrix
determinant calculation at every step of an MCMC, a new GP solver, {\tt
celerite} \citep{Foremanmackey2017} shows promise for increasing the
computional speed of GP rotation period inference by several orders of
magnitude.
We intend to integrate {\tt celerite} into our code in the near future.
How much computational time is too much computational time is a question that
depends on the available resources of each individual and their scientific
priorities.
GP period inference is trivially parallelisable, so if one has access to
thousands of computer cores, one can infer rotation periods of thousands of
stars in a few hours (perhaps just a few minutes with {\tt celerite}).
For single stars GP rotation inference takes a reasonable amount of time and
has already been used on exoplanet hosts \citep[\eg][]{Haywood2014,
Vanderburg2015}.

Is it worth spending the extra computation power to get slightly more
accurate, probabilistic rotation periods for a large number of stars?
A growing number of astronomers are interested in hierarchical probabilistic
modelling, particularly in the stellar and exoplanet communities.
These individuals are mindful of carefully inferring probabilistic parameters
wherever possible.
Probabilistic inference allows one to marginalise over nuisance
parameters while formally taking any uncertainty into account when
measuring the quantity of interest -- the star's rotation period, in
this case.
This becomes especially important as the signal-to-noise ratio
decreases and the rotation period is no longer tightly constrained,
and as the sampling becomes uneven.
One of our goals is to perform population inferences with a catalog of
rotation periods.
Throughout the astronomical literature, hierarchical Bayesian
inference has been demonstrated to be a robust method of population
inference \citep[\eg][]{Hogg2010, Foremanmackey2015, Wolfgang2015, Rogers2015}.
Probabilistic measurements, with full rigorous propagation of
uncertainties, are a crucial ingredient for any hierarchical inference
where the target is the population level distribution.
Inferring the ages of stars in the galaxy via gyrochronology (age-rotation
relations) \citep[\eg][]{Skumanich1972, Kawaler1989, Barnes2003, Barnes2007}
is an example of a hierarchical inference problem using stellar rotation.
Hierarchical inference allows you to marginalise over individual rotation
periods and infer age distribution parameters directly from the light curves.
Probabilistic rotation period inference is highly relevant to a large number
of galactic archaeology and exoplanet population studies which require
rotation period (or age) distributions.}

The code used in this project is available at
\url{https://github.com/RuthAngus/GProtation/} and samples of the KOI rotation
period posterior PDFs can be found at
\url{https://doi.org/10.5281/zenodo.804440}.
